---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/figures/README-"
)
```

# gsloid: global sea level and oxygen isotope data

The goal of gsloid is to make available raw data for sea level curves and $\delta$^18^O curves for the Holocene and most of the Pleistocene. 

## Installation

You can install gsloid from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("benmarwick/gsloid")
```

## Overview

This package includes two commonly used datasets in palaeoecology and archaeology:

- A global sea level curve
- A global oxygen isotope curve

There are many possible sources for these kinds of data, this package includes data from:

```{r, echo = FALSE}
references <- data.frame(Dataset = c("lisiecki2005", "spratt2016"),
                         Source = c("Lisiecki, L.E. and M.E. Raymo. 2005. A Pliocene-Pleistocene stack of 57 globally distributed benthic D18O records. Paleoceanography, Vol. 20, PA1003, doi:10.1029/2004PA001071.",
                                    "Spratt, Rachel M. and Lorraine E. Lisiecki 2016. A Late Pleistocene sea level stack. Climate of the Past. Vol. 12, 1079-1092, doi:10.5194/cp-12-1-2016"))

knitr::kable(references)
```

Here's the structure of the main datasets:

```{r}
library(gslic)
str(lisiecki2005)
```

```{r}
str(spratt2016)
```

Detailed descriptions of the variables are avaliable in the data documentation, run `?lisiecki2005` and `?spratt2016` for more information.

## Usage

Atlhough these data are suitable for many kinds of analyses, the primary reason that I made this package is so I can make my own plots of these data without having to copy a plot from another publication. Here's how I typically start with plotting the oxygen isotope data:

```{r}
library(ggplot2)


ggplot(lisiecki2005, 
       aes(Time,
           d18O)) +
  geom_line() +
  scale_x_continuous(limits = c(0, 250),
                     name = "x 1000 years ago") +
  scale_y_reverse(name = bquote(delta^18*O)) +
  theme_bw()
```

And here's how I start to plot the sea level data:

```{r}
ggplot(spratt2016, 
       aes(age_calkaBP,
           SeaLev_shortPC1)) +
  geom_line() +
  scale_x_continuous(limits = c(0, 250),
                     name = "x 1000 years ago") +
  scale_y_continuous(name = "Sea Level, meters above present day") +
  theme_bw()
```

Often we want to see the Marine istope stages on these plots also. This package includes the dataset `LR04_MISboundaries` of start and end dates for each stage, so we can draw these stages easily. 

Some care is required to get the MIS numbers positions in an easy-to-read location. Here I use `rep()` and `seq()` to help position the numbers. 

```{r}
# subset the MIS data for the last 250 ka years
mis_last_250ka <- LR04_MISboundaries[LR04_MISboundaries$LR04_Age_ka <= 250, ]

# plot the oxygen istope data line on top of the MIS lines
ggplot() +
  geom_vline(data = mis_last_250ka,          # add MIS lines
             aes(xintercept = LR04_Age_ka),
             colour = "blue") +
  annotate("text", 
           label = mis_last_250ka$label, 
           x = mis_last_250ka$LR04_Age_ka - 4,
           y = c(rep(3.0, 4), 
                 seq(3.2, 2.7, length.out = 6), 
                 rep(3.0, 2)),
           size = 3) +
  geom_line(data = lisiecki2005,                # add d18O line
            aes(Time,
                d18O),
            size = 0.75) +  
  scale_x_continuous(limits = c(0, 250),
                     name = "x 1000 years ago") +
  scale_y_reverse(name = bquote(delta^18*O)) +
  theme_bw()
```

Sometimes we prefer to indicate the MIS by horizontal lines. Once again we have to carefully place the line segments and labels so they are clear to read:

```{r}
ggplot() +
  geom_segment(data = mis_last_250ka, # add MIS lines
               aes(x =    line_start,
                   xend = LR04_Age_ka,
                   y =    seq(3, 2.5, length.out = nrow(mis_last_250ka)),
                   yend = seq(3, 2.5, length.out = nrow(mis_last_250ka))),
               colour = "blue",
               size = 1) +
  annotate("text", 
           label = mis_last_250ka$label, 
           x =     mis_last_250ka$mid,
           y = c(seq(2.9, 2.7, length.out = 4), 
                 seq(3.1, 2.8, length.out = 5),
                 seq(2.4, 2.3, length.out = 3)),
           size = 3) +
  geom_line(data = lisiecki2005,                # add d18O line
            aes(Time,
                d18O),
            size = 0.75) +  
  scale_x_continuous(limits = c(0, 250),
                     name = "x 1000 years ago") +
  scale_y_reverse(name = bquote(delta^18*O)) +
  theme_bw()
```

And sometimes we might want shaded rectangles to indicate the MIS:

```{r}
ggplot() +
  annotate("rect", 
           xmin = mis_last_250ka$line_start, 
           xmax = mis_last_250ka$LR04_Age_ka, 
           ymin = -Inf, 
           ymax = Inf,
        alpha = .2,
        fill = rep(c("grey50", "white"), 
                   nrow(mis_last_250ka)/2)) +
  annotate("text", 
           label = mis_last_250ka$label, 
           x =     mis_last_250ka$mid,
           y = c(rep(3, 4), 
                 seq(3.1, 2.8, length.out = 5),
                 rep(3, 3)),
           size = 3) +
  geom_line(data = lisiecki2005,                # add d18O line
            aes(Time,
                d18O),
            size = 0.75) +  
  scale_x_continuous(limits = c(0, 250),
                     name = "x 1000 years ago") +
  scale_y_reverse(name = bquote(delta^18*O)) +
  theme_bw()
```
